---
title: "Machine Learning Approaches to Dissect Hybrid and Vaccine-Induced Immunity"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Import Required Libraries

```{r message=FALSE, warning=FALSE}
# Load necessary libraries
library(mclust)  
library(umap)   
library(Rtsne)   
library(ggplot2) 
library(fpc)
library(caret)   
library(dplyr)   
```

## **Part 1: Dimensionality Reduction and Clustering**

#### Load and Explore Data

```{r}
set.seed(983) 
# Load the synthetic dataset
load("../data/synth1.rda")
str(synth1)  
summary(synth1)
```

#### **UMAP for Dimensionality Reduction**

```{r message=FALSE, warning=FALSE}
# Apply UMAP on the dataset, excluding the first column 
# (assumed to be Infectious status)
set.seed(289) 
um <- umap::umap(synth1[,-c(1)], method = 'umap-learn', preserve.seed = T)

# Convert UMAP results into a data frame
umap_df <- data.frame(um$layout)
umap_df$Group <- synth1$Infection_0_1pre_2post
umap_df$ID <- rownames(umap_df) 

# Plot UMAP projection
ggplot(umap_df, aes(x = X1, y = X2, color = Group)) + 
  geom_point(size = 3) +
  scale_color_manual(labels = c("No Infection", "Infection pre-boost", 
                                "Infection post-boost"),
                     values = c("#66a182", "#2e4057", "#edae49")) +
  labs(x = 'UMAP 1', y = 'UMAP 2', title = 'Dimensionality Reduction (UMAP)') +
  theme_minimal()
```

#### **Clustering with Gaussian Mixture Model (GMM)**

```{r}
# Clustering with Gaussian Mixture Model (GMM) on UMAP-reduced data (unsupervised)
um_gmm = mclust::Mclust(umap_df[, c(1, 2)]) 

# Print summary of clustering results
summary(um_gmm)

# Plot the clustering results
plot(um_gmm, "classification") 
plot(um_gmm, "density")
```

#### **t-SNE for Dimensionality Reduction**

```{r}
# Apply t-SNE on the dataset, excluding the first column 
# (assumed to be Infectious status)
set.seed(7567)
tsne <- Rtsne::Rtsne(synth1[, -c(1)], perplexity = 38, normalize=FALSE, theta = 0.5)

# Convert t-SNE results into a data frame
tsne_df <- data.frame(tsne$Y)
tsne_df$Group <- synth1$Infection_0_1pre_2post
tsne_df$ID <- rownames(tsne_df)

# Plot t-SNE projection
ggplot(tsne_df, aes(x = X1, y = X2, color = Group)) +
  geom_point(size = 3) +
  scale_color_manual(labels = c("No Infection", "Infection pre-boost", 
                                "Infection post-boost"),
                     values = c("#66a182", "#2e4057", "#edae49")) +
  labs(x = 't-SNE 1', y = 't-SNE 2', title = 'Dimensionality Reduction (t-SNE)') +
  theme_minimal()
```

#### **Clustering with Gaussian Mixture Model (GMM)**

```{r}
# Clustering with Gaussian Mixture Model (GMM) on t-SNE-reduced data (unsupervised)
t_gmm <- mclust::Mclust(tsne_df[, c(1, 2)])

# Print summary of clustering results
summary(t_gmm)

# Plot the clustering results
plot(t_gmm, "classification")
plot(t_gmm, "density")
```

#### **Clustering Evaluation: Within-cluster Sum of Squares & Silhouette Score**

```{r}
# Compute clustering statistics for UMAP-based clustering
cs_um_gmm <- fpc::cluster.stats(dist(umap_df[1:2]), um_gmm$classification)
stats_um_gmm <- cs_um_gmm[c("within.cluster.ss","avg.silwidth")]

# Compute clustering statistics for t-SNE-based clustering
cs_ts_gmm <- fpc::cluster.stats(dist(tsne_df[1:2]), t_gmm$classification)
stats_t_gmm <- cs_ts_gmm[c("within.cluster.ss","avg.silwidth")]

# Combine statistics and print Comparison
stats <- rbind(stats_um_gmm, stats_t_gmm)
rownames(stats) <- c("UMAP", "t-SNE")
stats <- as.data.frame(stats)
stats
```

------------------------------------------------------------------------

## **Part 2: Unaware Infection Prediction (Classification Task)**

#### **Load and Preprocess Data**

```{r}
# Load the labeled synthetic dataset (synth2)
load("../data/synth2.rda")

# Ensure the target variable is a factor
synth2$Class <- as.factor(synth2$Class)

# Display basic dataset structure and summary
str(synth2)
summary(synth2)
```

#### **Machine Learning Models Construction**

```{r}
# Set a global seed
set.seed(1058) 

# Create 5 cross-validation folds to use for all models
cv_folds <- caret::createFolds(synth2$Class, k = 5, returnTrain = TRUE)

# Define training control settings
control <- caret::trainControl(
  method = "cv", number = 5, 
  classProbs = TRUE, 
  summaryFunction = multiClassSummary, 
  verboseIter = TRUE, 
  index = cv_folds
)
```

```{r}
# Define a function for hyper parameter tuning and training models
train_model <- function(method, tL) {
  caret::train(Class ~ ., data = synth2, method = method, 
               trControl = control, tuneLength = tL)
}

# Train models with hyperparameter tuning
knn_model <- train_model("knn", tL = 5)
rf_model  <- train_model("rf", tL = 5)
svm_model <- train_model("svmRadial", tL = 5)
```

#### **Model Performance Evaluation**

```{r}
# Collect and compare model results
results <- resamples(list(kNN = knn_model, RF = rf_model, SVM = svm_model))

# Select metrics of interest
selected_metrics <- results$values %>% 
  select(contains(c("Accuracy", "Precision", "Recall", "F1")))

# Display performance summary
summary(selected_metrics)

# Visualization of model performance
bwplot(results)
dotplot(results)
```

#### **Prediction on Unlabeled Data**

```{r}
# Load synthetic unlabeled data (synth3)
load("../data/synth3.rda")

# Ensure feature consistency between synth2 and synth3
common_features <- intersect(names(synth2), names(synth3))
synth3 <- synth3[, common_features, drop = FALSE]

# Make predictions using trained models
synth3$kNN <- predict(knn_model, synth3)
synth3$RF  <- predict(rf_model, synth3)
synth3$SVM <- predict(svm_model, synth3)

# Compare model predictions
table(synth3$kNN)
table(synth3$RF)
table(synth3$SVM)
```
